---
title: "2020 Election Early Voting Summaries and Demographics"
author: "Aaron Kessler"
date: "10/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(lubridate)
library(readxl)
library(writexl)
library(tidycensus)
library(kableExtra)
options(scipen = 999)
options(dplyr.summarise.inform=F)


#run processing script to refresh with latest data as needed each day
# source("02_analysis.R")

# alternatively, use saved versions of files already created (much faster when new updated data not needed)
state_bothcycles <- readRDS("processed_data/state_bothcycles.rds")
# county_bothcycles <- readRDS("processed_data/county_bothcycles.rds")
natl_grandtots_bothyears <- readRDS("processed_data/natl_grandtots_bothyears.rds")
state_grandtots_bothyears <- readRDS("processed_data/state_grandtots_bothyears.rds")
# county_grandtots_bothyears <- readRDS("processed_data/county_grandtots_bothyears.rds")

# load lookup file with key state designations for prez and senate
keystates_lookup <- read_excel("processed_data/keystates_elex2020.xlsx")


#pull national votes cast/return number for use below
value_national_2016 <- natl_grandtots_bothyears %>% pull(x2016)
value_national_2020 <- natl_grandtots_bothyears %>% pull(x2020)
value_national_pctchg <- natl_grandtots_bothyears %>% pull(pctchg_returned) %>% round_half_up(1)

#we'll use use prez battlegrounds, create a vector to filer on
keystates_prez_vector <- keystates_lookup %>% 
  filter(office == "P") %>% 
  distinct(state) %>% 
  pull()

#use those values to create a subset of the states table
state_keyprezonly_bothcycles <- state_bothcycles %>% 
  filter(state %in% keystates_prez_vector)

state_keyprezonly_bothcycles


#we'll use use senate key races, create a vector to filer on
keystates_senate_vector <- keystates_lookup %>% 
  filter(office == "S") %>% 
  distinct(state) %>% 
  pull()

#use those values to create a subset of the states table
state_keysenateonly_bothcycles <- state_bothcycles %>% 
  filter(state %in% keystates_senate_vector)

state_keysenateonly_bothcycles



#### CHOOSE WHAT FILTERED DATASET TO USE FOR ANALYSIS
#select which filtered version of the full dataset to use below, or create new one
votedata <- state_keyprezonly_bothcycles


```

Bloomberg is partnering with Catalist, a data company that works with Democrats and others, to compile counts of ballots cast before Election Day, either early in-person or by-mail. The data is being provided each weekday morning.

As of the morning of `r format(Sys.Date(), "%B %d, %Y")`, at least **`r format(value_national_2020, big.mark=",")`** votes have been cast early or by mail nationwide, according to data collected by Catalist. This represents a **`r format(value_national_pctchg)`** percent change from 2016, when `r format(value_national_2016, big.mark=",")` votes were cast at this point.  
  
Details on ballots cast in selected states follow:

## Battleground State Totals

```{r, echo=FALSE}

state_grandtots_bothyears %>% 
  filter(state %in% keystates_prez_vector) %>% 
  select(state,
    "ballots cast at this point 2016" = total_returned_2016,
    "ballots cast to date 2020" = total_returned_2016,
    difference = diff_returned,
    "percent change" = pctchg_returned) %>% 
  kable(format.args = list(big.mark = ",")) %>% #formatting adds (,) 1000s separator
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                full_width = F,
                position = "left") %>% 
  column_spec(1, bold = T)


```


## Demographic breakdowns

### Gender

```{r, echo=FALSE}

#calculate growth for each gender group
gender_states <- votedata %>% 
  group_by(cycle, state, gender) %>% 
  summarise(
    total_returned = sum(ballots_returned, na.rm = TRUE)
  ) %>% 
  pivot_wider(names_from = c(cycle, gender), values_from = total_returned) %>% 
  clean_names() %>% 
  mutate(
    diff_female = x2020_female - x2016_female,
    pctchg_female = round_half_up((x2020_female - x2016_female) / x2016_female * 100, 2),
    diff_male = x2020_male - x2016_male,
    pctchg_male = round_half_up((x2020_male - x2016_male) / x2016_male * 100, 2),
    diff_unknown = x2020_unknown - x2016_unknown,
    pctchg_unknown = round_half_up((x2020_unknown - x2016_unknown) / x2016_unknown * 100, 2)
  ) 

#add totals for each cycle and pct totals
gender_states <- gender_states %>% 
  mutate(
    x2016_total = x2016_female + x2016_male + x2016_unknown,
    x2020_total = x2020_female + x2020_male + x2020_unknown,
    x2016_pcttotal_female = round_half_up(x2016_female / x2016_total * 100, 2),
    x2020_pcttotal_female = round_half_up(x2020_female / x2020_total * 100, 2),
    x2016_pcttotal_male = round_half_up(x2016_male / x2016_total * 100, 2),
    x2020_pcttotal_male = round_half_up(x2020_male / x2020_total * 100, 2),
    x2016_pcttotal_unknown = round_half_up(x2016_unknown / x2016_total * 100, 2),
    x2020_pcttotal_unknown = round_half_up(x2020_unknown / x2020_total * 100, 2),
    pct_chg_disparity_male_female = abs(pctchg_female - pctchg_male)
  )

# head(gender_states)

```


Growth in male vs. female early voters 

```{r, echo=FALSE}

#for displaying
gender_states %>% 
  select(
    state,
    "difference from 2016 in ballots cast by female voters" = diff_female,
    "difference from 2016 in ballots cast by male voters" = diff_male,
    "pct change from 2016 in ballots cast by female voters" = pctchg_female,
    "pct change from 2016 in ballots cast by male voters" = pctchg_male
  ) %>% 
  kable(format.args = list(big.mark = ",")) %>% #formatting adds (,) 1000s separator
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                full_width = F,
                position = "left") %>% 
  column_spec(1, bold = T)


```


### Age Groups


```{r, echo=FALSE}

# votedata

age_states <- votedata %>% 
  group_by(cycle, state, age) %>% 
  summarise(
    total_returned = sum(ballots_returned, na.rm = TRUE)
  ) %>% 
  pivot_wider(names_from = c(cycle), values_from = total_returned) %>% 
  clean_names() %>% 
  mutate(
    diff = x2020 - x2016,
    pctchg = round_half_up((x2020 - x2016) / x2016 * 100, 2)
  ) 

# age_states  
  
```


Growth in early voters by age groups to date, compared with the same point before the 2016 election

```{r, echo=FALSE}

#for display 
age_states %>% 
  select(
    state,
    age,
    "ballots cast at this point in 2016" = x2016,
    "ballots cast so far in 2020" = x2020,
    "difference" = diff,
    "pct change" = pctchg
  ) %>% 
  kable(format.args = list(big.mark = ",")) %>% #formatting adds (,) 1000s separator
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                full_width = F,
                position = "left") %>% 
  column_spec(1, bold = T)


```




```{r, echo=FALSE}




```


```{r, echo=FALSE}




```


```{r, echo=FALSE}




```


```{r, echo=FALSE}




```


```{r, echo=FALSE}




```


```{r, echo=FALSE}




```

